---
title: "Thesis -version2"
author: "Farida Eleshin"
date: "2/8/2022"
output:
  word_document: default
  html_document: default
  pdf_document: default
---


##IMPORTING DATA SET
```{r}
#test data - predictor variables
test_set_features_flu = read.csv("/Users/macpro/Desktop/Data_/test_set_features_flu.csv", header = T) 


#training data - predictor variables
training_set_features_flu = read.csv("/Users/macpro/Desktop/Data_/training_set_features_flu.csv", header = T)


#response variables
training_set_labels_flu = read.csv("/Users/macpro/Desktop/Data_/training_set_labels_flu.csv", header = T)

```

   
####Checking for missing values in training dataset dataset
```{r}
sapply(training_set_features_flu, function(x)sum(is.na(x)))
```

#checking for missing values in the test dataset
```{r}
sapply(test_set_features_flu, function(x)sum(is.na(x)))
```

#checking for missing values in the labels dataset
```{r}
sapply(training_set_labels_flu, function(x)sum(is.na(x)))
```

#renaming datasets
```{r}
trainingdata<-training_set_features_flu
testdata <- test_set_features_flu
response.labels<-training_set_labels_flu
```


##Changing all categorical data to numeric data and creating new columns of them
```{r}
trainingdata[is.na(trainingdata)] = 0

trainingdata$age_group <-as.factor(trainingdata$age_group)
trainingdata$newage <- as.numeric(trainingdata$age_group)

trainingdata$race <-as.factor(trainingdata$race)
trainingdata$newrace <- as.numeric(trainingdata$race)

trainingdata$education <-as.factor(trainingdata$education)
trainingdata$neweducation <- as.numeric(trainingdata$education)

trainingdata$sex <-as.factor(trainingdata$sex)
trainingdata$newsex <- as.numeric(trainingdata$sex)

trainingdata$income_poverty <-as.factor(trainingdata$income_poverty)
trainingdata$newincome_poverty <- as.numeric(trainingdata$income_poverty)

trainingdata$marital_status <-as.factor(trainingdata$marital_status)
trainingdata$newmarital_status <- as.numeric(trainingdata$marital_status)

trainingdata$rent_or_own <-as.factor(trainingdata$rent_or_own)
trainingdata$newrent_or_own <- as.numeric(trainingdata$rent_or_own)

trainingdata$employment_status <-as.factor(trainingdata$employment_status)
trainingdata$newemployment_status <- as.numeric(trainingdata$employment_status)

trainingdata$hhs_geo_region <-as.factor(trainingdata$hhs_geo_region)
trainingdata$newhhs_geo_region <- as.numeric(trainingdata$hhs_geo_region)

trainingdata$census_msa <-as.factor(trainingdata$census_msa)
trainingdata$newcensus_msa <- as.numeric(trainingdata$census_msa)

trainingdata$employment_industry <-as.factor(trainingdata$employment_industry)
trainingdata$newemployment_industry <- as.numeric(trainingdata$employment_industry)

trainingdata$employment_occupation <-as.factor(trainingdata$employment_occupation)
trainingdata$newemployment_occupation <- as.numeric(trainingdata$employment_occupation)




```
#Deleting old categorical variables
```{r}

trainingdata[, 23:32] <- list(NULL)
trainingdata[, 35:36] <- list(NULL)

```

##Changing all categorical data to numeric data and creating new columns of them - test data set
```{r}
testdata[is.na(testdata)] = 0

testdata$age_group <-as.factor(testdata$age_group)
testdata$newage <- as.numeric(testdata$age_group)

testdata$race <-as.factor(testdata$race)
testdata$newrace <- as.numeric(testdata$race)

testdata$education <-as.factor(testdata$education)
testdata$neweducation <- as.numeric(testdata$education)

testdata$sex <-as.factor(testdata$sex)
testdata$newsex <- as.numeric(testdata$sex)

testdata$income_poverty <-as.factor(testdata$income_poverty)
testdata$newincome_poverty <- as.numeric(testdata$income_poverty)

testdata$marital_status <-as.factor(testdata$marital_status)
testdata$newmarital_status <- as.numeric(testdata$marital_status)

testdata$rent_or_own <-as.factor(testdata$rent_or_own)
testdata$newrent_or_own <- as.numeric(testdata$rent_or_own)

testdata$employment_status <-as.factor(testdata$employment_status)
testdata$newemployment_status <- as.numeric(testdata$employment_status)

testdata$hhs_geo_region <-as.factor(testdata$hhs_geo_region)
testdata$newhhs_geo_region <- as.numeric(testdata$hhs_geo_region)

testdata$census_msa <-as.factor(testdata$census_msa)
testdata$newcensus_msa <- as.numeric(testdata$census_msa)

testdata$employment_industry <-as.factor(testdata$employment_industry)
testdata$newemployment_industry <- as.numeric(testdata$employment_industry)

testdata$employment_occupation <-as.factor(testdata$employment_occupation)
testdata$newemployment_occupation <- as.numeric(testdata$employment_occupation)

```

#Deleting old categorical variables
```{r}

testdata[, 23:32] <- list(NULL)
testdata[, 35:36] <- list(NULL)


```

# # ##removing duplicated respondent ID numbers
```{r}
# # ##removing duplicated ID numbers
 trainingdata <- trainingdata[!duplicated(trainingdata$respondent_id), ]
testdata <- testdata[!duplicated(testdata$respondent_id), ]
testdata[-c(26708), ]

```

##replacing na values with 0
```{r}
trainingdata[is.na(trainingdata)]  = 0
testdata[is.na(testdata)]  = 0
```


#LEAST ABSOLUTE SHRINKAGE AND SELECTION OPERATOR (LASSO) REGRESSION AS A PREDICTION MODEL 
```{r}
#H1N1 VACCINE AS THE RESPONSE VARIABLE

#install.packages(glmnet)

#loading package glmnet
library(glmnet)

#assigning x to the training data and y to h1n1_vaccine in the response labels
x.h1n1<-trainingdata
y.h1n1 <- response.labels$h1n1_vaccine

#fitting the first lasso model 
fit<-glmnet(x.h1n1,y.h1n1)

#Visualizing the coefficients
plot(fit)
#CV =cv.glmnet(x,y)
```

```{r}
print(fit)

```

#Making predictions on the test data using the first lasso model generated
```{r}
#Making predictions  
set.seed(101)
test.data.matrix <- data.matrix(testdata)
lasso.predictions <-predict(fit, newx = test.data.matrix, s =1.0)

#printing the first 20
head(lasso.predictions)
```
At this specific lambda, the probability that everyone has taken h1n1 is the same. Which doesn't explain the individual different probability we need so we do a cross validation

######Cross Validation
```{r}
x2.h1n1 <-data.matrix(trainingdata)
cvfit <- cv.glmnet(x2.h1n1, y.h1n1, family = "binomial", type.measure = "auc", 
                  keep = TRUE)
plot(cvfit)
```


```{r}
#the minimum lambda - lambda.min returns the value of lambda that gives minimum mean cross-validated error.
cvfit$lambda.min

```



####Finding the coefficient of lambda 1se - most regularised model
```{r}
cvfit$lambda.1se
coef(cvfit, s = "lambda.1se")
```

#Predicting the probabilities of the varibles on h1n1 vaccine using lambda.lse, which gives the most regularized model such that error is within one standard error of the minimum.
```{r}
#lambda.lse, which gives the most regularized model such that error is within one standard error of the minimum.
lasso.bestpred <- predict(cvfit, newx = test.data.matrix, s = "lambda.1se")
head(lasso.bestpred)
```


#Assessing using the confusion matrix
```{r}
cnf <- confusion.glmnet(cvfit, newx = x2.h1n1, newy = y.h1n1)
print(cnf)
```
#Assessing using the ROC curve
```{r}
rocs <- roc.glmnet(cvfit$fit.preval, newy = y.h1n1)
best <- cvfit$index["min",]
plot(rocs[[best]], type = "l")
invisible(sapply(rocs, lines, col="grey"))
lines(rocs[[best]], lwd = 2,col = "red")
```

#SEASONAL VACCINE - LASSO REGRESSION
#Training the data for seasonal vaccine
```{r}
#install.packages(glmnet)
library(glmnet)

x.seasonal.vaccine<-trainingdata
y.seasonal.vaccine <- response.labels$seasonal_vaccine

fit<-glmnet(x.seasonal.vaccine,y.seasonal.vaccine)
plot(fit)
#print(fit)


```

#Making predictions using the test data
```{r}
#Making predictions at the specific lambda 
set.seed(101)
nx <- data.matrix(testdata)
pred <-predict(fit, newx = nx, s =1.0)

#printing the first 20
head(pred)
```


######Cross Validation
```{r}
x.seasonal.vaccine.data.matrix <-data.matrix(trainingdata)
cvfit <- cv.glmnet(x.seasonal.vaccine.data.matrix, y.seasonal.vaccine,family = "binomial", type.measure = "auc", 
                  keep = TRUE)
plot(cvfit)
```


#VIWEING THE SELECTED LAMBDAS AND THEIR CORRESPONDING COEEFICIENTS
```{r}
#the minimum lambda - lambda.min returns the value of lambda that gives minimum mean cross-validated error.
cvfit$lambda.min
```


####Finding the coefficient of lambda lse
```{r}
cvfit$lambda.1se
coef(cvfit, s = "lambda.1se")
```

#Predicting the probabilities of the varibles on seasonal vaccine using lambda.lse, which gives the most regularized model such that error is within one standard error of the minimum.
```{r}
#lambda.lse, which gives the most regularized model such that error is within one standard error of the minimum.
x.seasonal.vaccine.testdata.data.matrix<-data.matrix(testdata)
bestpred <- predict(cvfit, newx = x.seasonal.vaccine.testdata.data.matrix, s = "lambda.1se")
head(bestpred)
```

#Assessing using the confusion matrix
```{r}
cnf <- confusion.glmnet(cvfit, newx = x2.h1n1, newy = y.seasonal.vaccine)
print(cnf)
```

#Assessing using the roc curve
```{r}
rocs <- roc.glmnet(cvfit$fit.preval, newy = y.seasonal.vaccine)
best <- cvfit$index["min",]
plot(rocs[[best]], type = "l")
invisible(sapply(rocs, lines, col="grey"))
lines(rocs[[best]], lwd = 2,col = "red")
```


#ARTIFICIAL NEURAL NETWORK AS A PREDICTION MODEL 
```{r}
#install.packages("neuralnet")

```

#H1N1 ANN MODEL

##training data set
```{r}
#training <-trainingdata
#test <- testdata

df=data.frame(trainingdata,response.labels$h1n1_vaccine)
head(df,5)

# h1n1<-training_set_labels_flu$h1n1_vaccine
```

##Creating the neural network
```{r}
# load library
require(neuralnet)


# fit neural network
nn = neuralnet(response.labels.h1n1_vaccine~h1n1_concern+h1n1_knowledge+`behavioral_antiviral_meds`+`behavioral_avoidance`+`behavioral_face_mask`+`behavioral_wash_hands`+`behavioral_large_gatherings`+`behavioral_outside_home`+`behavioral_touch_face`+`doctor_recc_h1n1`+`doctor_recc_seasonal`+`chronic_med_condition`+`child_under_6_months`+`health_worker`+`health_insurance`+`opinion_h1n1_vacc_effective`+`opinion_h1n1_risk`+`opinion_h1n1_sick_from_vacc`+`opinion_seas_vacc_effective`+`opinion_seas_risk`+`opinion_seas_sick_from_vacc`+`newage`+`newrace`+`neweducation`+`newsex`+`newincome_poverty`+`newmarital_status`+`newrent_or_own`+`newemployment_status`+`newemployment_industry`+`newemployment_occupation`,data=df, hidden=1,act.fct = "logistic",
                linear.output = FALSE)
plot(nn)

nn$result.matrix



```


```{r}
## Prediction using neural network
#testdata<-testdata[-c(2),]
#test <- testdata
#testdata[is.na(testdata)]  = 0



print("Probabilties that each each respondent took an h1n1 vaccine")
nn.h1n1.predictions<-predict (nn,testdata)
head(nn.h1n1.predictions,5)


print("H1n1 probabilties converted into binary")
nn.binary <- ifelse(nn.h1n1.predictions>0.5, 1, 0)
head(nn.binary,10)
```

#Cross validation of the neuralnetwork
```{r}
#library(caret)
#nn.binary<-nn.binary[-c(2),]
#nn.binarydf<-data.frame(nn.binary,df$response.labels.h1n1_vaccine)
#attach(nn.binarydf)
table(nn.binarydf)
conf <-(18042+823)/(18042+4851+ 2991+823)
conf

```


###FOR SEASONAL VACCINE - ANN MODEL 
```{r}
#training <-trainingdata
#test <- testdata
#training[is.na(training)] = 0



df1=data.frame(trainingdata,response.labels$seasonal_vaccine)
#head(df,20)

require(neuralnet)



nn = neuralnet(response.labels.seasonal_vaccine~h1n1_concern+h1n1_knowledge+`behavioral_antiviral_meds`+`behavioral_avoidance`+`behavioral_face_mask`+`behavioral_wash_hands`+`behavioral_large_gatherings`+`behavioral_outside_home`+`behavioral_touch_face`+`doctor_recc_h1n1`+`doctor_recc_seasonal`+`chronic_med_condition`+`child_under_6_months`+`health_worker`+`health_insurance`+`opinion_h1n1_vacc_effective`+`opinion_h1n1_risk`+`opinion_h1n1_sick_from_vacc`+`opinion_seas_vacc_effective`+`opinion_seas_risk`+`opinion_seas_sick_from_vacc`+`newage`+`newrace`+`neweducation`+`newsex`+`newincome_poverty`+`newmarital_status`+`newrent_or_own`+`newemployment_status`+`newemployment_industry`+`newemployment_occupation`,data=df1, hidden=1,act.fct = "logistic",
 
                             linear.output = FALSE)

#plotting the neural network 
plot(nn)
nn$result.matrix

#changing all na values to na
#test[is.na(test)]  = 0

#predict using test data
nn.seasonalvaccine.predictions<-predict (nn,testdata)

print("H1n1 probabilties converted into binary")
nn.sv.binary <- ifelse(nn.seasonalvaccine.predictions>0.5, 1, 0)
head(nn.sv.binary,10)

```



```{r}
nn.sv.binary<-nn.sv.binary[-c(2),]
nn.sv.binarydf <-data.frame(nn.sv.binary,response.labels$seasonal_vaccine)
attach(nn.sv.binarydf)
table(nn.sv.binarydf)
conf.ss<-(8157+5490)/(8157+5490+6115+6945)
conf.ss
```

#XGBOOST 
```{r}
library(xgboost)
library(caTools)
library(dplyr)
#library(caret)

testdata<-testdata[-c(2),]

###data
xgb_train <- xgb.DMatrix(data = data.matrix(trainingdata), label =df$response.labels.h1n1_vaccine )
xgb_test <- xgb.DMatrix(data = data.matrix(testdata), label = df$response.labels.h1n1_vaccine)

set.seed(42)
xgb_params <- list(
  booster = "gbtree",
  eta = 0.01,
  max_depth = 8,
  gamma = 4,
  subsample = 0.75,
  colsample_bytree = 1,
  objective = "binary:logistic",
  eval_metric = "auc",
  num_class = 0
)


xgb_model <- xgb.train(
  params = xgb_params,
  data = xgb_train,
  nrounds = 5000,
  verbose = 1
)
xgb_model

xgb_preds <- predict(xgb_model, data.matrix(testdata))
xgb_preds <- data.frame(xgb_preds)
#colnames(xgb_preds) <- levels(response.labels$h1n1_vaccine)
xgb_preds


```


```{r}
library(caret)
xgb_preds.bn <- ifelse(xgb_preds>0.5, 1, 0)
#xgb_preds.bn
accuracy <- sum(xgb_preds.bn == df$response.labels.h1n1_vaccine) / nrow(xgb_preds.bn)
accuracy

confusionMatrix(factor(xgb_preds.bn), factor(df$response.labels.h1n1_vaccine))


```
#XGBOOST SEASONAL VACCINE
```{r}
library(xgboost)
library(caTools)
library(dplyr)
library(caret)

xgb_train <- xgb.DMatrix(data = data.matrix(trainingdata), label =df1$response.labels.seasonal_vaccine )
xgb_test <- xgb.DMatrix(data = data.matrix(testdata), label = df1$response.labels.seasonal_vaccine)

set.seed(42)
xgb_params <- list(
  booster = "gbtree",
  eta = 0.01,
  max_depth = 8,
  gamma = 4,
  subsample = 0.75,
  colsample_bytree = 1,
  objective = "binary:logistic",
  eval_metric = "auc",
  num_class = 0
)


xgb_model <- xgb.train(
  params = xgb_params,
  data = xgb_train,
  nrounds = 5000,
  verbose = 1
)
xgb_model

xgb_preds <- predict(xgb_model, data.matrix(testdata))
xgb_preds <- data.frame(xgb_preds)
xgb_preds

xgb_preds.bn <- ifelse(xgb_preds>0.5, 1, 0)
#xgb_preds.bn
accuracy <- sum(xgb_preds.bn == df1$response.labels.seasonal_vaccine) / nrow(xgb_preds.bn)
accuracy

confusionMatrix(factor(xgb_preds.bn), factor(df1$response.labels.seasonal_vaccine))

```



#GRADIENT BOOSTING MACHINE
```{r}
library(caret)        
library(scales)
library(tidyverse)
library(gbm)
library(Metrics)
library(here)

#trainingdata[, 25] <- list(NULL)
# trainingdata[, 26] <- list(NULL)
#testdata[, 25] <- list(NULL)
# testdata[, 26] <- list(NULL)
```


```{r}
Boston.boost.h1n1=gbm(response.labels$h1n1_vaccine ~ . ,data = trainingdata,distribution = "gaussian",n.trees = 10000,
                  shrinkage = 0.01, interaction.depth = 4)
Boston.boost.h1n1

summary(Boston.boost.h1n1) #Summary gives a table of Variable Importance and a plot of Variable Importance
```
```{r}
#Plot of Response variable with doctor_recc_h1n1
plot(Boston.boost.h1n1,i="doctor_recc_h1n1") 


plot(Boston.boost.h1n1,i="health_insurance") 

```
```{r}
cor(trainingdata$doctor_recc_h1n1,response.labels$h1n1_vaccine)#negetive correlation coeff-r

cor(trainingdata$health_insurance,response.labels$h1n1_vaccine)#positive correlation coeff-r
```

```{r}
n.trees = seq(from=100,to=10000,by=100) #no of trees-a vector of 100 values 

#Generating a Prediction matrix for each Tree
predmatrix<-predict(Boston.boost.h1n1,testdata,n.trees=n.trees)
dim(predmatrix) #dimensions of the Prediction Matrix 

#Calculating The Mean squared Test Error
test.error<-with(testdata,apply( (predmatrix-response.labels$h1n1_vaccine)^2,2,mean))
head(test.error) #contains the Mean squared test error for each of the 100 trees averaged

#Plotting the test error vs number of trees

plot(n.trees , test.error , pch=19,col="blue",xlab="Number of Trees",ylab="Test Error", main = "Perfomance of Boosting on Test Set")

#adding the RandomForests Minimum Error line trained on same data and similar parameters
abline(h = min(test.error),col="red") #test.err is the test error of a Random forest fitted on same data
legend("topright",c("Minimum Test error Line for Random Forests"),col="red",lty=1,lwd=1)
```
```{r}
#predtrial<-predmatrix
#newms<-predtrial[, 4:100] <- list(NULL)
#gbm.binary<-ifelse(predmatrix>0.5, 1, 0)
# conf.h<-confusionMatrix(factor(gbm.binary), factor(response.labels$h1n1_vaccine)) 
# conf.h
accuracy.gbm <- sum(gbm.binary == df$response.labels.h1n1_vaccine) / nrow(gbm.binary)
accuracy.gbm

```

#GBM FOR SEASONAL VACCINE
```{r}
library(caret)        
library(scales)
library(tidyverse)
library(gbm)
library(Metrics)
library(here)

Boston.boost.sv=gbm(response.labels$seasonal_vaccine ~ . ,data = trainingdata,distribution = "gaussian",n.trees = 10000,
                  shrinkage = 0.01, interaction.depth = 4)
Boston.boost.sv

summary(Boston.boost.sv) 

#Plot of Response variable with doctor_recc_h1n1
plot(Boston.boost.sv,i="doctor_recc_h1n1") 
#Inverse relation with lstat variable

plot(Boston.boost.sv,i="health_insurance") 
#as the average number of rooms increases the the price increases

cor(trainingdata$doctor_recc_h1n1,response.labels$seasonal_vaccine)#negetive correlation coeff-r
cor(trainingdata$health_insurance,response.labels$seasonal_vaccine)#positive correlation coeff-r

n.trees = seq(from=100,to=10000,by=100) #no of trees-a vector of 100 values 

#Generating a Prediction matrix for each Tree
predmatrix<-predict(Boston.boost.sv,testdata,n.trees=n.trees)
dim(predmatrix) #dimensions of the Prediction Matrix 

#Calculating The Mean squared Test Error
test.error<-with(testdata,apply( (predmatrix-response.labels$seasonal_vaccine)^2,2,mean))
head(test.error) #contains the Mean squared test error for each of the 100 trees averaged

#Plotting the test error vs number of trees

plot(n.trees , test.error , pch=19,col="blue",xlab="Number of Trees",ylab="Test Error", main = "Perfomance of Boosting on Test Set")

#adding the RandomForests Minimum Error line trained on same data and similar parameters
abline(h = min(test.error),col="red") #test.err is the test error of a Random forest fitted on same data
legend("topright",c("Minimum Test error Line for Random Forests"),col="red",lty=1,lwd=1)

gbm.binary.sv<-ifelse(predmatrix>0.5, 1, 0)

accuracy.gbm <- sum(gbm.binary.sv == response.labels$seasonal_vaccine) / nrow(gbm.binary.sv)
accuracy.gbm
```


